### 2017-04-27 paper/jfp-submission/evaluation.tex
tweak citations

\begin{figure}[htp]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.5\linewidth]{distrib.png}
\end{minipage}
}
\vspace{3ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 79--85\% of
  the programs in under one second, improving to 87\% in under
  10 seconds. In both datasets we detect actual type errors about 82\%
  of the time, unbound variables or constructors 3--4\% of the time, and
  diverging loops 2--3\% of the time. For the remaining 11--12\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for 83\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we hit a maximum of 87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 3--4\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 2--3\% triggered an infinite
recursion error.
%
For the remaining 12\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely trigger more of the type errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

\paragraph{Example: Recursion with Bad Base Case}
%
The function @sumList@ should add up
the elements of its input list.
%
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> ==[]==
    | y::ys -> y + __sumList ys__
\end{ecode}
%
Unfortunately, in the base case, it returns @[]@
instead of @0@.
%
\sherrloc\ blames the base case, and \ocaml\
assumes the base case is correct and blames
the \emph{recursive call} on line 3:
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
Both of the above are parts of the full story, which
is summarized by \toolname's trace showing
how @sumList [1; 2]@ gets stuck at @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
The trace clarifies (via the third step)
that the @[]@ is the result of the recursive call
@sumList []@, and shows how it is incompatible with
the subsequent \texttt{+} operation.

@@

  let rec digitsOfInt n =
    if n <= 0 then
      []
    else
      append (==digitsOfInt (n / 10)==)
             [__n mod 10__]
\end{ecode}
%
%\pagebreak
Unfortunately, the student's @append@ function \emph{conses} an element
onto a list instead of appending two lists.
%
Though incorrect, @append@ still type-checks and thus \ocaml and
\sherrloc blame the \emph{use-site} on line 10.
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
In contrast, \toolname makes no assumptions about @append@
yielding a trace that illustrates the true error on line 4, by
highlighting the conflict in consing a list onto a list of integers.
%
\begin{center}
  \includegraphics[height=160px]{digitsOfInt.png}
\end{center}
%

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors: (1) with high coverage in a timespan amenable to
compile-time analysis; (2) with traces that have a low average
complexity of 7 jumps; (3) that are more helpful to novice
programmers than traditional type error messages; and (4) that
can be used to automatically locate the source of a type error.

There are, of course, drawbacks to our approach. Four that stand out
are: (1) coverage limits due to random generation, (2) the inability to
handle certain instances of infinite types, (3) dealing with an
explosion in the size of generated traces, and (4) handling ad-hoc
polymorphism.

\paragraph{Random Generation}
Random test generation has difficulty generating highly constrained
values, \eg\ red-black trees or a pair of equal integers. If the type
error is hidden behind a complex branch condition \nanomaly\ may not be
able to trigger it. Exhaustive testing and dynamic-symbolic execution
can address this short-coming by performing an exhaustive search for
inputs (\emph{resp}.\ paths through the program). As our experiments
show, however, novice programs do not appear to require more advanced
search techniques, likely because the novice programs tend to be simple.

>>>>

\begin{figure}[htp]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.5\linewidth]{distrib.png}
\end{minipage}
}
\vspace{3ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 75--82\% of
  the programs in under one second, improving to 84\% in under
  10 seconds. In both datasets we detect actual type errors at least 75\%
  of the time, unbound variables or constructors 4--5\% of the time, and
  diverging loops 3\% of the time. For the remaining 13--16\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for at least 75\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we hit a maximum of 83--86\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 4--5\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 3\% triggered an infinite
recursion error.
%
For the remaining 13--16\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely uncover more errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

\paragraph{Example: Recursion with Bad Base Case}
%
The function @sumList@ should add up
the elements of its input list.
%
\begin{ecode}
  let rec sumList xs = match xs with
    | []    -> ==[]==
    | y::ys -> y + __sumList ys__
\end{ecode}
%
Unfortunately, in the base case, it returns @[]@
instead of @0@.
%
\sherrloc\ blames the base case, and \ocaml\
assumes the base case is correct and blames
the \emph{recursive call} on line 3:
%
\begin{verbatim}
  This expression has type
    'a list
  but an expression was expected of type
    int
\end{verbatim}
%
Both of the above are parts of the full story, which
is summarized by \toolname's trace showing
how @sumList [1; 2]@ gets stuck at @2 + []@.
%
\begin{center}
  \includegraphics[height=125px]{sumlist.png}
\end{center}
%
The trace clarifies (via the third step)
that the @[]@ results from the recursive call
\hbox{@sumList []@,} and shows how it is incompatible with
the subsequent \texttt{+} operation.

@@

  let rec digitsOfInt n =
    if n <= 0 then
      []
    else
      append (==digitsOfInt (n / 10)==)
             [__n mod 10__]
\end{ecode}
%
%\pagebreak
Unfortunately, the student's @append@ function \emph{conses} an element
onto a list instead of appending two lists.
%
Though incorrect, @append@ still type-checks and thus \ocaml and
\sherrloc blame the \emph{use-site} on line 10.
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
In contrast, \toolname makes no assumptions about @append@,
yielding a trace that illustrates the error on line 4, by
highlighting the conflict in consing a list onto a list of integers.
%
\begin{center}
  \includegraphics[height=160px]{digitsOfInt.png}
\end{center}
%

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors:
%
(1) with high coverage in a timespan amenable to compile-time analysis;
%
(2) with traces that have a low average complexity of 7 jumps;
%
(3) that are more helpful to novice programmers than traditional type
error messages; and
%
(4) that can be used to automatically locate the source of a type error.

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) the inability to handle certain instances of infinite types;
%
(3) dealing with an explosion in the size of generated traces; and
%
(4) handling ad-hoc polymorphism.

\paragraph{Random Generation}
Random test generation has difficulty generating highly constrained
values, \eg\ red-black trees or a pair of equal integers. If the type
error is hidden behind a complex branch condition \nanomaly\ may not be
able to trigger it. Exhaustive testing and dynamic-symbolic execution
can address this short-coming by performing an exhaustive search for
inputs (\emph{resp}.\ paths through the program). As our experiments
show, however, novice programs do not appear to require more advanced
search techniques, likely because they tend to be simple.

### 2017-04-27 paper/jfp-submission/related.tex
tweak citations

\paragraph{Localizing and Repairing Type Errors}
\label{sec:diagnosis-repair}
% It is well known that
% unification-based type inference procedures can
% produce poor error messages, and in particular, can misidentify the
% \emph{source} of the type error.
%
Many groups have explored techniques to pinpoint the true source
of errors reported by static type checkers.
% unification-based type inference procedures.
%
% \subparagraph{Localization}
The traditional Damas-Milner type inference algorithm~\cite{Damas1982-uw}
reports the first program location where a type mismatch is discovered
(subject to the traversal strategy~\cite{Lee1998-ys}).
%
As a result the error can be reported far away from its
source~\cite{McAdam1998-ub} without enough information to guide the
user.
%
Type-error slicing~\cite{Haack2003-vc,Schilling2011-yf,Rahli2015-tt,Sagonas2013-bf,Gast2004-zd,Neubauer2003-xv}
%treats type inference as a constraint-satisfaction problem and
recognizes this flaw and instead produces a slice of the program
containing \emph{all} program locations that are connected to the type
error.
%
%% \cite{Neubauer2003-xv} present a decidable type system based on
%% discriminative sum types, in which all terms are typeable and type
%% derivations contain all type errors in a program. They then use the
%% typing derivation to slice out the parts of the expression related to
%% each error.
%
Though the program slice must contain the source of the error, it can
suffer from the opposite problem of providing \emph{too much}
information, motivating recent work in ranking the candidate locations.
%
Zhang~\etal~\cite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
identifying the most likely culprit using Bayesian reasoning.
%
\cite{Pavlinovic2014-mr,Pavlinovic2015-kh} translate the %error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.

% \subparagraph{Repairing}
In addition to localizing the error, \cite{Lerner2007-dt} attempt to
suggest a fix by replacing expressions (or removing them entirely) with
alternatives based on the surrounding program context.
%
\cite{Chen2014-gd} use a variational type system to allow for the
possibility of changing an expression's type, and search for an
expression whose type can be changed such that type inference would
succeed.
%
% It then attempts to deduce the error source by searching for an
% expression whose type can be changed such that type inference would
% succeed.
%
In contrast to Lerner~\etal, who search for changes at the value-level,
Chen~\etal\ search at the type-level and are thus complete due the finite
universe of types used in the program.
%
% \item \cite{chen_error-tolerant_2012}
% \item \cite{okeefe_type_1992}
% \item \cite{gomard_partial_1990}
% \item \cite{thatte_type_1988}
%

In contrast to these approaches, we do not attempt to localize or fix
the type error. Instead we try to explain it to the user using a
dynamic witness that demonstrates how the program is not just
ill-typed but truly wrong. In addition, allowing users to run their
program % (even knowing that it is wrong)
enables experimentation and the
use of debuggers to step through the program and investigate its
evolution.

\paragraph{Improving Error Messages}
%
The content and quality of the error messages themselves has also been
studied extensively.
%
\cite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
messages in novice environments and present suggestions for improving
their quality and consistency.
%
\cite{Hage2006-hc} identify a variety of general heuristics to improve
the quality of type error messages, based on their teaching experience.
%
\cite{Heeren2003-db}, \cite{Christiansen2014-qc}, and
\cite{Serrano2016-oo} provide methods for library authors to specialize
type errors with domain-specific knowledge.
%
The difference with our work is more pronounced here as we do not
attempt to improve the quality of the error message, instead we search
for a witness to the error and explain it with the resulting execution
trace.
%

\paragraph{Running Ill-Typed Programs}
\label{sec:running-ill-typed}
\cite{Vytiniotis2012-gh} extend the \haskell
compiler GHC to support compiling ill-typed programs, but their intent
is rather different from ours. Their goal was to allow programmers to
incrementally test refactorings, which often cause type errors in
distant functions. They replace any expression that fails to type
check with a \emph{runtime} error, but do not check types
at runtime.
%
\cite{Bayne2011-cn} also provide a semantics for running
ill-typed (\java) programs, but in constrast transform the program to
perform nearly all type checking at run-time. The key difference between
Bayne~\etal\ and our work is that we use the dynamic semantics to
automatically search for a witness to the type error, while their focus
is on incremental, programmer-driven testing.

@@

Our operational semantics is amenable to dynamic-symbolic execution, one
would just need to collect the path condition and replace our
implementation of \gensym by a call to the solver. We chose to use lazy,
random generation instead because it is efficient, and % does not incur
the overhead of an external solver, and produces high coverage for our
domain of novice programs.

A function's type is a theorem about the function's behavior.
Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
thereby connecting it to work on using test generation to find
counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.

@@

Flanagan \etal~\cite{Flanagan96} describe a static debugger for Scheme, which helps
the programmer interactively visualize problematic source-sink flows
corresponding to soft-typing errors. The debugger allows the user to explore
an abstract reduction graph computed from a static value set analysis of
the program. In contrast, \toolname generates witnesses and allows the user
to explore the resulting dynamic execution.
%
Perera \etal~\cite{Perera2012-dy} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} --- containing holes --- and present a partial slice,
containing only those steps that affected the the partial value.
% This
% system is designed to answer questions of the form ``Where did this
% value come from?'' and thus is focused on backward exploration.
Perera~\etal\ focus on backward exploration; in contrast, our
visualization supports forward \emph{and} backward exploration, though
our backward steps are more limited.
%
Specifically, we do not support selecting a value and inserting the
intermediate terms that preceded it while ignoring unrelated computation
steps. %; this would be interesting future work.

>>>>

\paragraph{Localizing and Repairing Type Errors}
\label{sec:diagnosis-repair}
% It is well known that
% unification-based type inference procedures can
% produce poor error messages, and in particular, can misidentify the
% \emph{source} of the type error.
%
Many groups have explored techniques to pinpoint the true source
of errors reported by static type checkers.
% unification-based type inference procedures.
%
% \subparagraph{Localization}
The traditional Damas-Milner type inference algorithm~\cite{Damas1982-uw}
reports the first program location where a type mismatch is discovered
(subject to the traversal strategy~\cite{Lee1998-ys}).
%
As a result the error can be reported far away from its
source~\cite{McAdam1998-ub} without enough information to guide the
user.
%
Type-error slicing~\cite{Haack2003-vc,Schilling2011-yf,Rahli2015-tt,Sagonas2013-bf,Gast2004-zd,Neubauer2003-xv}
%treats type inference as a constraint-satisfaction problem and
recognizes this flaw and instead produces a slice of the program
containing \emph{all} program locations that are connected to the type
error.
%
%% \cite{Neubauer2003-xv} present a decidable type system based on
%% discriminative sum types, in which all terms are typeable and type
%% derivations contain all type errors in a program. They then use the
%% typing derivation to slice out the parts of the expression related to
%% each error.
%
Though the program slice must contain the source of the error, it can
suffer from the opposite problem of providing \emph{too much}
information, motivating recent work in ranking the candidate locations.
%
Zhang~\etal~\shortcite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
identifying the most likely culprit using Bayesian reasoning.
%
Pavlinovic~\etal~\shortcite{Pavlinovic2014-mr,Pavlinovic2015-kh}
translate the %error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.
%
Loncaric~\etal~\shortcite{Loncaric2016-uk} improve the scalability of
Pavlinovic~\etal by reusing the existing type checker as
a theory solver in the Nelson-Oppen~\shortcite{Nelson1979-td}
style, thus requiring only a MaxSAT solver.

% \subparagraph{Repairing}
In addition to localizing the error, Lerner~\etal~\shortcite{Lerner2007-dt} attempt to
suggest a fix by replacing expressions (or removing them entirely) with
alternatives based on the surrounding program context.
%
Chen~\&~Erwig~\shortcite{Chen2014-gd} use a variational type system to allow for the
possibility of changing an expression's type, and search for an
expression whose type can be changed such that type inference would
succeed.
%
% It then attempts to deduce the error source by searching for an
% expression whose type can be changed such that type inference would
% succeed.
%
In contrast to Lerner~\etal, who search for changes at the value-level,
Chen~\&~Erwig search at the type-level and are thus complete due the finite
universe of types used in the program.
%
% \item \cite{chen_error-tolerant_2012}
% \item \cite{okeefe_type_1992}
% \item \cite{gomard_partial_1990}
% \item \cite{thatte_type_1988}
%

In contrast to these approaches, we do not attempt to localize or fix
the type error. Instead we try to explain it to the user using a
dynamic witness that demonstrates how the program is not just
ill-typed but truly wrong. In addition, allowing users to run their
program (even knowing that it is wrong)
enables experimentation and the
use of debuggers to step through the program and investigate its
evolution.

\paragraph{Improving Error Messages}
%
The content and quality of the error messages themselves has also been
studied extensively.
%
Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
messages in novice environments and present suggestions for improving
their quality and consistency.
%
Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general heuristics to improve
the quality of type error messages, based on their teaching experience.
%
Heeren~\etal~\shortcite{Heeren2003-db},
Christiansen~\shortcite{Christiansen2014-qc}, and
Serrano~\&~Hage~\shortcite{Serrano2016-oo}
provide methods for library authors to specialize
type errors with domain-specific knowledge.
%
The difference with our work is more pronounced here as we do not
attempt to improve the quality of the error message, instead we search
for a witness to the error and explain it with the resulting execution
trace.
%

\paragraph{Running Ill-Typed Programs}
\label{sec:running-ill-typed}
Vytiniotis~\etal~\shortcite{Vytiniotis2012-gh} extend the \haskell
compiler GHC to support compiling ill-typed programs, but their intent
is rather different from ours. Their goal was to allow programmers to
incrementally test refactorings, which often cause type errors in
distant functions. They replace any expression that fails to type
check with a \emph{runtime} error, but do not check types
at runtime.
%
Bayne~\etal~\shortcite{Bayne2011-cn} also provide a semantics for running
ill-typed (\java) programs, but in constrast transform the program to
perform nearly all type checking at run-time. The key difference between
Bayne~\etal\ and our work is that we use the dynamic semantics to
automatically search for a witness to the type error, while their focus
is on incremental, programmer-driven testing.

@@

Our operational semantics is amenable to dynamic-symbolic execution, one
would just need to collect the path condition and replace our
implementation of \gensym by a call to the solver. We chose to use lazy,
random generation instead because it is efficient, does not incur
the overhead of an external solver, and produces high coverage for our
domain of novice programs.

A function's type is a theorem about the its behavior.
Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
thereby connecting it to work on using test cases to find
counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.

@@

Flanagan~\etal~\shortcite{Flanagan96} describe a static debugger for Scheme, which helps
the programmer interactively visualize problematic source-sink flows
corresponding to soft-typing errors. The debugger allows the user to explore
an abstract reduction graph computed from a static value set analysis of
the program. In contrast, \toolname generates witnesses and allows the user
to explore the resulting dynamic execution.
%
Perera~\etal~\shortcite{Perera2012-dy} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} --- containing holes --- and present a partial slice,
containing only those steps that affected the the partial value.
% This
% system is designed to answer questions of the form ``Where did this
% value come from?'' and thus is focused on backward exploration.
Perera~\etal\ focus on backward exploration; in contrast, our
visualization supports forward \emph{and} backward exploration, though
our backward steps are more limited.
%
Specifically, we do not support selecting a value and inserting the
intermediate terms that preceded it while ignoring unrelated computation
steps. %; this would be interesting future work.

### 2017-04-27 paper/jfp-submission/safe.tex
tweak citations

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We used a combination of automatic and manual coding to categorize these
programs into four classes.
%
The first class is easily detected by \toolname itself, and thus admits
a precise count.
%
This left us with 473 programs requiring manual coding; we selected
a random sample of 50 programs to investigate, and will report results
based on that sample.
%
Figure~\ref{fig:no-witness} summarizes the results of our investigation,
we note the classes that were based on a random sample with a ``*''.

@@

\paragraph{Overloaded Functions.}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound, we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in @bad@, from type errors that were caused by our instantiation.

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, we can still view @x@ and @y@
\emph{morally} as holes, and thus treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

@@

\paragraph{Non-Parametric Function Type *.}
%
5\% of programs lack a witness in our semantics due to our
non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code and ``Safe'' Function Calls *.}
%
4\% of programs contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

1\% contained type errors that were unreachable by any inputs, often due
to overlapping patterns in a @match@ expression.
%
% While technically safe, dead code is generally considered a ``code
% smell''; students would likely benefit from a warning in this case.
% \ES{Wes says this is too informal, source the claim}
%
While technically safe, dead code is generally considered a maintenance
risk, as the programmer may not realize that it is dead~\cite{Wheeler2014-fg}
or may accidentally bring it back to life~\cite{Seven2014-gf}.
%
Thus, a warning like that provided by \ocaml's pattern-exhaustiveness
checker would be helpful.

%\paragraph{Bad Function Calls}
A further 3\% included a function call where the
student had supplied ill-typed inputs, yet the path induced
by the call contained no errors.
%
Consider, for example the following |assoc| function, which
looks up a key in an \emph{association list} and returns a default if
the key cannot be found.
%
\begin{code}
  let rec assoc (d, k, l) = match l with
    | (ki, vi)::tl ->
       if ki = k then
         vi
       else
         assoc (d, k, tl)
    | _ -> d

  let _ = assoc ([], 123, [(123, "sad"); (321, "happy")])
\end{code}
%
The student's definition of |assoc| is correct, but \ocaml rejects their
subsequent call because the default value |[]| is incompatible with the
|string| values in the list.
%
In this particular call, the key |123| is in the list, so the default
will not be used (and even if it were, no error would be triggered) and
\ocaml's complaint is moot.

Of course, \ocaml cannot be expected to know that this particular call
is safe, its type system is not sophisticated enough to express the
necessary conditions.
%
% \ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}

@@

\paragraph{Witness Exists *.}
%
We found that only 2\% of programs admit a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.

\paragraph{Summary.}
%
Our investigation suggests that the majority of programs
(84\%) for which we fail to find a witness do not, in fact, admit a
witness.
%
Overall, it appears that our random instantiation of holes is well-suited
to finding type errors in student programs.
%
The remaining programs were all cases where \ocaml's type system was
overly conservative.
%
Of course, the conservatism is somewhat justified as each case pointed
to code that would be difficult to use or maintain;
%
it would be interesting to investigate how demonstrate these issues in a
similar manner.

>>>>

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We used a combination of automatic and manual coding to categorize these
programs into four classes.
%
The first class is easily detected by \toolname itself, and thus admits
a precise count.
%
This left us with 473 programs that required manual coding; we selected
a random sample of 50 programs to investigate, and will report results
based on that sample.
%
Figure~\ref{fig:no-witness} summarizes the results of our investigation ---
we note the classes that were based on the random sample with a ``*''.

@@

\paragraph{Overloaded Functions}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound; we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in @bad@, from type errors that were caused by our instantiation.

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given \pagebreak
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, we can still view @x@ and @y@
\emph{morally} as holes, and thus treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

@@

\paragraph{Non-Parametric Function Type *}
%
5\% of programs lack a witness in our semantics due to our
non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code and ``Safe'' Function Calls *}
%
4\% of programs contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

1\% contained type errors that were unreachable by any inputs, often due
to overlapping patterns in a @match@ expression.
%
% While technically safe, dead code is generally considered a ``code
% smell''; students would likely benefit from a warning in this case.
% \ES{Wes says this is too informal, source the claim}
%
While technically safe, dead code is generally considered a maintenance
risk, as the programmer may not realize that it is dead~\cite{Wheeler2014-fg}
or may accidentally bring it back to life~\cite{Seven2014-gf}.
%
Thus, a warning like that provided by \ocaml's pattern exhaustiveness
checker would be helpful.

%\paragraph{Bad Function Calls}
A further 3\% included a function call where the
student supplied ill-typed inputs, but the path induced
by the call did not contain an error.
%
Consider the following |assoc| function, which
looks up a key in an \emph{association list}, returning a default if
it cannot be found.
%
\begin{code}
  let rec assoc (d, k, l) = match l with
    | (ki, vi)::tl ->
       if ki = k then
         vi
       else
         assoc (d, k, tl)
    | _ -> d

  let _ = assoc ([], 123, [(123, "sad"); (321, "happy")])
\end{code}
%
The student's definition of |assoc| is correct, but \ocaml rejects their
subsequent call because the default value |[]| is incompatible with the
|string| values in the list.
%
In this particular call the key |123| is in the list, so the default
will not be used (even if it were, there would not be an error) and
\ocaml's complaint is moot.
%
Of course, \ocaml cannot be expected to know that this particular call
is safe, its type system is not sophisticated enough to express the
necessary conditions.
%
% \ES{TODO: add some comment about using dependent types, though this particular example would be quite bizarre to encode in DT..}

@@

\paragraph{Witness Exists *}
%
We found that only 2\% of programs admit a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.

\paragraph{Summary}
%
Our investigation suggests that the vast majority of programs
for which we fail to find a witness do not, in fact, admit a
witness.
%
% Overall, it appears that our random instantiation of holes is well-suited
% to finding type errors in student programs.
These programs were generally cases where \ocaml's type system was
overly conservative.
%
Of course, the conservatism is somewhat justified as each case pointed
to code that would be difficult to use or maintain;
%
it would be interesting to investigate how demonstrate these issues in an
intuitive manner.
